Tax-Equity-in-Low-and-Middle-Income-Countries-Replication

### 2024-02-19 paper/paper.qmd
Finish the findings section in Discussion

This paper expands upon the analysis presented in "Tax Equity in Low- and Middle-Income Countries" by Bachas, Jensen, and Gadenne, focusing on how economic growth shapes tax systems and influences tax equity [@article]. By examining the transition from indirect taxation, such as VAT, to direct taxation, including personal income taxes, as economies develop, this research provides insights into the mechanisms through which tax systems evolve and their implications for income inequality.

@@

1. Shift from Indirect to Direct Taxation: Our analysis, supported by data visualization in @fig-taxes-gdp, illustrates a clear trend of countries moving towards a greater reliance on direct taxes as their GDP per capita increases. This shift signifies a crucial step towards more progressive tax systems that can potentially reduce income inequality by placing a larger tax burden on those with higher incomes.

@@

The research methodology and the analysis of the datasets have allowed us to draw these conclusions and offer a nuanced understanding of tax systems' evolution with economic development. By employing the R programming language and the tidyverse package for data cleaning and analysis, we have meticulously examined the role of taxation in economic contexts characterized by varying degrees of informality and enforcement capabilities.

>>>>

This paper expands upon the analysis presented in "Tax Equity in Low- and Middle-Income Countries" by Bachas, Jensen, and Gadenne @article, focusing on how economic growth shapes tax systems and influences tax equity. By examining the transition from indirect taxation, such as VAT, to direct taxation, including personal income taxes, as economies develop, this research provides insights into the mechanisms through which tax systems evolve and their implications for income inequality.

@@

1. Shift from Indirect to Direct Taxation: Our analysis, supported by data visualization in @fig-taxes-gdp and @fig-taxrevenue-gdp-diff-source, illustrates a clear trend of countries moving towards a greater reliance on direct taxes as their GDP per capita increases. This shift signifies a crucial step towards more progressive tax systems that can potentially reduce income inequality by placing a larger tax burden on those with higher incomes.

@@

A more detailed discussion on each finding will be provided in the following sections.

## Shift from Indirect to Direct Taxation

The dynamics of tax policy are fundamentally intertwined with economic development, influencing not only fiscal stability but also social equity within nations. Taxation, encompassing both direct and indirect levies, serves as a crucial mechanism for redistributing income, funding public services, and shaping economic incentives. The relationship between tax structures and economic growth has been a subject of extensive study, revealing a nuanced interplay that varies significantly across different stages of economic development.

A pivotal aspect of this relationship is the observed shift from indirect to direct taxation as economies mature. Indirect taxes, such as sales taxes or VAT, are applied uniformly to transactions irrespective of the buyer's income level, potentially placing a disproportionate burden on lower-income households. Conversely, direct taxes, including personal and corporate income taxes, can be more effectively tailored to the payer's ability to pay, allowing for a more progressive taxation system. This shift not only reflects a country's evolving economic landscape but also its commitment to fostering social equity through tax policy.

### Empirical Evidence and Global Comparisons

The empirical evidence from @fig-taxes-gdp and @fig-taxrevenue-gdp-diff-source in the study vividly illustrates the transition from indirect to direct taxation as economies advance. This shift is evidenced by the positive correlation between GDP per capita and the share of GDP collected through direct taxes, excluding social security contributions. The data showcases a clear trend: as countries' economies develop, there's a marked decrease in the reliance on indirect taxes such as VAT, sales taxes, and excises, with a concurrent uptick in direct taxes, particularly personal income taxes. This pattern suggests a move towards more progressive tax systems in higher GDP per capita economies, where tax policies increasingly lean on direct taxation mechanisms to foster equitable economic growth and redistribution.

This observation aligns with global trends noted in academic literature. For instance, Genschel and Seelkopf's @genschel2016tax research on taxation trends outside the OECD reveals that while non-Western countries adopt Western tax trends like increasing general revenues and implementing consumption taxes, they struggle with establishing a strong, progressive personal income tax system. This struggle is attributed to various socio-economic and structural factors, which, despite global influences, hinder the replication of progressive tax regimes seen in more developed economies.

Moreover, Stoilova and Patonov @stoilova2013empirical provide a comprehensive analysis on the impact of taxation on economic growth within the European Union, offering empirical evidence of how tax structures affect economic performance from 1995 to 2010. Their findings, alongside the examination of tax revenue trends and social contributions in European Union countries by Dubrovina et al. @dubrovina2019nexus, underscore the significance of adapting tax policies to enhance fiscal roles and ensure equitable distribution without overburdening the population.

These studies corroborate the trends depicted in the provided figures, reinforcing the global narrative of evolving tax structures as economies grow. They not only validate the empirical evidence from the study but also expand the understanding of how these trends manifest in different contexts, from OECD countries to emerging economies. Such analyses underscore the complexity and dynamism of tax policy evolution, reflecting a broad consensus on the direction of tax policy reforms aimed at fostering economic development and social equity.

### Tax Policy's Impact on Income Inequality

The shift towards direct taxation, as economies grow, has potential implications for reducing income inequality, as evidenced by @fig-avg-tax-rate-vs-income-decile. These shifts suggest that as countries' GDP per capita increases, there's a greater reliance on progressive taxation mechanisms, which are known for their potential to reduce income disparities by imposing a larger tax burden on higher-income earners.

Literature supports this potential, with studies highlighting the role of tax policy in addressing income inequality. Ballard-Rosa, Martin, and Scheve @ballardrosa2017structure delve into the political economy aspects, suggesting that preferences over taxation significantly influence policy responses to income inequality. Moreover, Możdżierz @mozdzierz2015financialisation identifies financialization as a primary driver of increasing income inequality, indicating the need for tax policies that can counteract these trends effectively.

Further, research by Liu Qiongzhi highlights the decisive role of policy bias in determining fiscal redistribution policies, emphasizing the importance of favoring equitable tax rates and transfer payments to address income disparities [@liu2020policybias]. Khan and Padda @khan2021taxexpenditures find that in South Asia, taxation and social expenditure policies effectively reduce income inequality, highlighting the effectiveness of direct taxes and higher social spending as tools for income redistribution.

The complexity and challenges in designing effective tax policies are underscored by outliers in data, suggesting that not all shifts towards direct taxation seamlessly lead to reduced income inequality. Factors such as tax evasion and the informal economy pose significant challenges. Studies like Esaku's @esaku2021incomeinequality demonstrate how income inequality can exacerbate the size of the shadow economy, further complicating tax policy design and implementation. Additionally, Wildowicz-Szumarska @wildowiczszumarska2022eu discusses the effectiveness of redistributive policies within the EU, indicating varied success across different welfare state models in tackling income inequality through tax policy.
